Experience Report: Haskell in Computational Biology

### 2012-06-20 mrfy_experience_report.tex
damnfino what's going on, but these are my changes

\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size

@@

\newenvironment{smallverbatim}{\par\small\verbatim}{\endverbatim}
\newcommand\smallverbatiminput[1]{{\par\unskip\small\verbatiminput{#1}}}
\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}

@@

MRFy searches for placements stochastically,
trying to minimize scores.
We~are exploring several stochastic algorithms: random-mutation hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
Our~exploration relies on higher-order functions.

@@

MRFy's search computes a sequence of \emph{generations}.
A~generation is a collection of scored {placements}.
We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
\ifnotcutting
It~too considers the list of accepted generations and the age of the current
generation. 
\fi
\end{itemize}
\seclabel{search}%
We hope you find the search simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}

The \texttt{SearchStrategy} interface provides ample scope for
experiments. 
Random-mutation hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a generation only if it scores better than its
predecessor;
simulated annealing may accept a generation that scores worse.)
Our genetic algorithm reuses the same
\texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextGen} function took forty lines of code
and a full day to implement.
\ifnotcutting
Our~current \texttt{quit} function simply runs for a fixed number of
generations, but its interface will enable us to try other criteria,
such as stopping once $k$~generations have passed without more than $E$
improvement in score. 
\fi

@@

We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford
Larsen, and Nathan Ricci. 
\fi

>>>>

\newif\ifnotcutting \notcuttingtrue % set if not cutting down to size

@@

\newenvironment{smallverbatim}{\par\small\verbatim}{\endverbatim}
\newcommand\smallverbatiminput[1]{{\par\unskip\small\verbatiminput{#1}}}
\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}

@@

MRFy searches for placements stochastically,
trying to minimize scores.
We~are exploring several stochastic algorithms: random-mutation hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
Our~exploration exploits higher-order functions, existentially
quantified types, and (to a degree) lazy evaluation.

@@

We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
of \emph{points}. 
Each point contains
a collection of {placements}; different searches represent
collections differently.
Each point also has a \texttt{Score}; \mrfy\ looks for points with
good scores.

Ideally, MRFy's search would use now-classic lazy, modular
technique advocated by \citet{hughes:why}, in which one function
computes an infinite sequence of points, and another function
uses a finite prefix to decide on an approximation.
But because MRFy's search is stochastic, 
making MRFy's search lazy and modular is not so easy.

To illustrate the difficulties, we~use our simplest search:
random hill climbing.
The space of beta-strand placements has no notion of slope or
gradient,
so at any given point in the search space, this search moves a random
distance in a random 
direction.
If~it leads to a better point, the move
is \emph{useful};
otherwise it is \emph{useless}.
\smallverbatiminput{utility}
(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
With luck, an~infinite sequence of useful moves converges
at a local optimum.

\mrfy's search path follows only useful moves;
if a move proves useless, 
\mrfy\ abandons it and tries again (with a new random seed) at the
previous point.
Ideally, \mrfy\ would search by composing a \emph{generator}
that produces an infinite sequence of moves, 
a \emph{filter} that selects the useful moves, and 
a \emph{test} function that enumerates finitely many useful moves
and returns the final destination.
But the generator may produce an infinite sequence of useless moves.
(For example, if \mrfy\ should find a global optimum, every move from
that point is guaranteed to be useless.)
Given an infinite sequence of useless inputs, a filter would not be
productive, and a test function would diverge.

We~address this problem by combining ``generate and filter'' into a
single abstraction, which has type \texttt{SearchGen pt r}.
Type variable~\texttt{pt} is a point in the search
space, and type variable~\texttt{r} is a
random-number generator:
\smallverbatiminput{gen.tex}
The monadic computation \texttt{pt0} randomly selects a starting point
for search;
\texttt{nextPt} produces a new point from
an existing point.
Because scoring may be expensive, both \texttt{pt0}
and \texttt{nextPt} use \emph{scored} points, and they can
reuse scores from previous points.

Why do we have \texttt{pt0} and \texttt{nextPt}
instead 
of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
which could produce an infinite list?
Because after the first move
the starting point for the next move depends on whether the first move
was useful.\remark{I hate this ``explanation.'' ---NR}
That decision is made by the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{age} of the \texttt{younger} point which is
moved~to.\remark{This word ``age'' is \emph{entirely} bogus!}
For example, in~simulated annealing, as the search progresses,
the \texttt{utility} function becomes less likely to express a move
that results in a worse score.\remark{And I hate this explanation
too.}

\noindent
\vbox{
Using these pieces, function \texttt{everyGen} produces an infinite
sequence containing a mix of useful and useless moves:
\smallverbatiminput{everygen}
}
\noindent
Because both \texttt{nextPt} and \texttt{utility} consume randomness,
the computation is very sequential, but it does use a little bit of
laziness:
from its starting point, \texttt{everyPt} provides an infinite list of
randomly chosen successor points, then calls \texttt{agedUtility} to
tag each one with its age and utility.\remark{There's that bogus age
again.}
The Prelude function \texttt{span} finds the first useful successor,
but in case there \emph{is} no successor, the code must also produce
all the useless successors.
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.

The most telling line of code in \texttt{everyPt} is the last line of
the body, which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with another useful point.
When the computation returned by \texttt{everyPt} is evaluated against
a random-number generator, it returns an infinite list containing a
mix of useful and useless points.
This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
\vbox{
\smallverbatiminput{strat}
\smallverbatiminputx{stop}
\smallverbatiminputx{search}
}
\noindent
We have implemented two kinds of stop functions...

N.B. WE HAVE TWO KINDS OF NON-MODULARITY.  FIRST, TERMINATION HAS TO
LOOK AT \emph{ALL} STATES, EVEN THOUGH WE WOULD PREFER THAT IT LOOK
ONLY AT USEFUL STATES.  SECOND, AGE IS EVERYWHERE.  IN PARTICULAR, WE
CANNOT SEPARATE THE AGE OF THE CURRENT STATE FROM THE UTILITY
DECISION.

\remark{We need to explain why we don't just generate an infinite lazy
list of thingies.  And perhaps we can consider an infinite tree!}

The \texttt{SearchStrategy} interface provides ample scope for
experiments.\remark{\P\ needs rewriting}
Random-mutation hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.
\ifnotcutting
Our~current \texttt{quit} function simply runs for a fixed number of
generations, but its interface will enable us to try other criteria,
such as stopping once $k$~generations have passed without more than $E$
improvement in score. 
\fi

\subsection*{The old search algorithm}

We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{orig-scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{orig-strategy}
In this record,
\begin{itemize}
\item
Function \texttt{pt0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
\ifnotcutting
It~too considers the list of accepted generations and the age of the current
generation. 
\fi
\end{itemize}
\seclabel{search}%
We hope you find the search simple and straightforward:
\smallfuzzverbatiminput{11pt}{orig-search}

@@

An~anonymous referee\remark{Just one referee?} spurred us to think
more deeply about laziness.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci. 
\fi

### 2012-06-20 search.tex
damnfino what's going on, but these are my changes

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Score])
search strat score (s:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s
  runFrom :: [Seed] -> [Scored Placement] -> [Score]
          -> Age -> (Scored Placement, [Score])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist = (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)

>>>>

search :: SearchStrategy pt r -> Rand r (History pt)
search (SS strat test) =
  fmap test . everyPt strat 0 =<< pt0 strat
